---
title: "Chapter_6_Time_Series"
author: "Daniel T. Citron"
date: "1/10/2020"
output: html_document
---

# Chapter 6: Time-Series Analysis

## Time domain: Autocorrelation and ARMA

ACF = autocorrelation function, quantifies serial correlations at different time lags

ARMA = Autoregressive moving average

Both of these are classic tools for describing serial dependence in time series in the *time domain*.

### ACF
Apply ACF to the weekly time series of prevalence from the seasonally forced SEIR model, for time lags up to 3 years.

Defining the SEIR model:
```{r}
library(deSolve)

seirmod2 <- function(t, y, parms){
  # vector of initial conditions
  S = y[1]
  E = y[2]
  I = y[3]
  R = y[4]
  # return the derivative
  with(as.list(parms), {
    dS = mu * (N - S) - beta0 * (1 + beta1 * cos(2 * pi * t)) * S* I/N
    dE = beta0 * ( 1 + beta1 * cos(2*pi*t)) * S * I/N - (mu + sigma)*E
    dI = sigma * E - (mu + gamma) * I
    dR = gamma * I - mu * R
    res = c(dS, dE, dI, dR)
    list(res)
  })
}
```

```{r}
times = seq(0,100, by = 1/52)
paras = c(mu = 1/50, N = 1, 
          beta0 = 1000, beta1 = 0.2, sigma= 365/8, gamma = 365/5)
xstart = c(S = 0.06, E = 0, I = 0.001, R = 0.939)
out = as.data.frame(ode(xstart, times, seirmod2, paras))

head(round(out,3))
```

```{r}
par(mfrow = c(1,2))
plot(times, out$I, ylab = "Infected", xlab = "Time", xlim = c(90, 100), type = "l")
acf(out$I, lag.max = 156, main = "")
```

## ARMA
Application: forecasting disease dynamics (ILI)

ARMA(p,q) assumes that the future incidence $Y_t$ can be predicted according to $$ Y_t = a_1 Y_{t - 1} + \dots + a_p Y_{t - p} + \varepsilon_t - b_1 \varepsilon_{t - 1} - \dots - b_q \varepsilon_{t - q} $$

$\varepsilon_t$ represents stochasticity, including the effects of past stochasticity

Load data from the flu in iceland, and convert to a time series `ts` object and do a seasonal decomposition. The `decompose` function uses moving averages to perform a Classical Seasonal Decomposition - splits the time series into a set of additive trends: The original series, seasonal component, the trend, and the residuals (randomness)
```{r}
library(epimdr)
library(forecast)
data(Icelandflu)

ilits <- ts(data = sqrt(Icelandflu$ili), 
            start = c(1980, 1), end = c(2009, 12), frequency = 12)
plot(decompose(ilits, type ="additive"))
```

We can now train a seasonal model for the years of 1990-2000, and perform a 24-month forecast
ARMA(2,1), meaning that we look back across 2 years (24 months), and refer to the past 1 year of noise
```{r}
wts = window(ilits, start = c(1990, 6), end = c(2000,5))
fit = arima(sqrt(wts), order = c(2, 0, 1),
            list(order = c(1,0,0), period=12))
coef(fit)
```
The forecast:
```{r}
fore = predict(fit, n.ahead = 24)

# upper and lower limits on the forecast:
U = fore$pred + 2 * fore$se
L = fore$pred - 2 * fore$se

ts.plot(sqrt(wts), fore$pred, U, L, col = c(1,2,4,4),
        lty = c(1,1,2,2), ylab = "Sqrt (cases)")
legend("bottomleft", c("ILI", "Forecast",
"95% Error Bounds"), col=c(1, 2, 4),lty=c(1, 1, 2))
```

## Frequency domain

### Schuster periodogram
A direct way of estimating and testing for (significant) periodicity - it decomposes a time series into cycles of different frequencies, and measures the importance weight of each frequency using spectral amplitude using an FFT.
```{r}
# spectrum() function calculates the periodogram
my.spec <- spectrum(out$I, plot = FALSE)

# make two plots:
par(mfrow = c(1,2))
# plot frequency
plot(my.spec, xlab="Frequency", ylab="Log-amplitude", main="", sub="")
# plot with period on the x axis
plot(1/my.spec$freq/52, my.spec$spec, type = "b", xlab = "Period (year)", ylab = "Amplitude", xlim = c(0,5))
```

### Wavelet analysis
An extension of spectral analysis - combines the time + frequency domains. There is no set of "canonical" periods for decomposition, but we can pick a specific wavelet which will dictate which sets of periods we need

Define a different SEIR model:
```{r}
seirmod = function(t, y, parms) { 
    S = y[1]
    E = y[2]
    I = y[3]
    R = y[4]
    mu = parms["mu"]
    N = parms["N"]
    beta = parms["beta"]
    sigma = parms["sigma"]
    gamma = parms["gamma"]
    dS = mu * (N - S) - beta * S * I/N
    dE = beta * S * I/N - (mu + sigma) * E 
    dI = sigma * E - (mu + gamma) * I
    dR = gamma * I - mu * R
    res = c(dS, dE, dI, dR)
    list(res)
}
```

```{r}
# Simulate the time series, and plot
times = seq(0, 25, by = 1/52)
paras = c(mu = 1/50, N = 1,
          beta = 1000,
          sigma = 365/8, gamma = 365/5)
xstart = c(S = 0.06, E = 0, I = 0.001, R = 0.939)
out2 = as.data.frame(ode(xstart, times, seirmod, paras))

plot(times, out2$I, type="l", xlab="Time",
     ylab = "Infected")
```

Perform the wavelet analysis - we need to specify the "octaves" and the number "voices" to perform the Morlet wavelet analysis (from the `Rwave` package). 8 octaves means the main period will be $\{2^1, \dots, 2^8\}$. The number of voices specifies the subdivisions of each octave. Four voices means the resultant periods will include $\{2^1, 2^{1.25}, 2^{1.75}, 2^2, \dots \}$
```{r}
#install.packages("Rwave")
require(Rwave)
# Set the number of "octaves" and "voices"
no = 8; nv = 32;
# calculating periods:
a = 2^seq(1, no + 1 - 1/nv, by = 1/nv)
# cwt function from the Rwave package
wfit = cwt(out2$I, no, nv, plot = FALSE)
wspec = Mod(wfit)
```

Plot
```{r}
par(mfrow = c(1, 2)) #Side-by-side plots
plot(times, out2$I, type="l", xlab="Time", ylab = "Infected")


image(x=times, wspec, col=gray((12:32)/32), y=a/52, ylim=c(0,4), xlab="Time", ylab="Period")
contour(x=times, wspec, y=a/52, ylim=c(0,4), zlim=c(mean(wspec), max(wspec)), add=TRUE)
```
How do we interpret this? At the start, the inter-epidemic period is strong, but then wanes as the system converges toward a stable epidemic (the amplitude of the oscillations decreases noticeably from the simulated data)

We can take slices through this contour plot and comapre the period spectrum amplitude at different years:
```{r}
plot(a/52, wspec[104,], type="l", ylab="Amplitude", xlab="Period")
lines(a/52, wspec[1040,], type="l", lty=2, col="red")
legend("topright", legend=c("Year 2", "Year 10"), lty=c(1,2), col=c("black", "red"))

```


## Measles in London
Pre-vaccination measles data from London during post-WW2

```{r}
data(meas)
head(meas)
# This includes case counts and birth counts on a 2-week interval from 1944 through 1964
```
Plot:
```{r}
par(mar = c(5,5,2,5)) #Make room for two axes 
plot(meas$time, meas$London, type="b", xlab="Week", ylab="Incidence", ylim=c(0,8000)) 
par(new=T) #Superimposed births plot 
plot(meas$time, meas$B, type="l", col="red", axes=FALSE, xlab=NA, ylab=NA, ylim=c(1000, 2700)) 
axis(side = 4)
mtext(side = 4, line = 3, "Births") 
legend("topright", legend=c("Cases", "Births"), lty=c(1,1), col=c("black", "red"))
```
We can apply wavelet analysis to the measles dynamics from London during this period (as per Grenfell 2001)

On top of the wavelet analysis, we will use the "crazy climber" algorithm to identify ridges in the wavelet function - highlighting significant features
```{r}
#Set octaves, voices and associated periods
no = 8; nv = 32
a = 2^seq(1,no+1-1/nv, by = 1/nv) #Continous wavelet decomposition
wfit = cwt(meas$London, no, nv, plot=FALSE) 
wspec = Mod(wfit)
#Crazy climber
crcinc<-crc(wspec, nbclimb=10, bstep=100)
fcrcinc<-cfamily(crcinc, ptile=0.5, nbchain=1000, bstep=10)
## There are 2 chains.
ridges<-fcrcinc[[1]]
ridges[which(ridges==0)]<-NA
#Wavelet plot with crazy-climber and contours 
image(x=meas$time, wspec, col=gray((12:32)/32), y=a/26, ylim=c(0.1,3), ylab="Period", xlab="Year")
contour(x=meas$time, wspec, y=a/26, ylim=c(0,3), nlevels = 6, zlim=c(mean(wspec), max(wspec)), add=TRUE)
image(x=meas$time, y=a/26, z=ridges, add=TRUE, col=gray(0))
```

We can again take slices through this, comparing 1945 with 1954: we see the transition from annual to biennial epidemics: this is predicted by seasonally forced SEIR + dropping birth rates
```{r}
plot(a/26,wspec[261,], type="l",xlim=c(0,3), xlab="period (years)", ylab="amplitude")
lines(a/26,wspec[27,], type="l", lty=2, col="red") 
legend("topleft", legend=c("1945", "1954"), lty=c(2,1), col=c("red", "black"))
```

