---
title: "R_0"
author: "Daniel T. Citron"
date: "1/9/2020"
output: html_document
---

# Chapter 3 - $$R_0$$

```{r}
library(epimdr)
```

Measles data from 2003 in Niger
```{r}
data("niamey")
head(niamey)
```

Plot cases over time: (Figure 3.1 in the text)
```{r}
par(mar = c(5,5,2,5))
plot(niamey$absweek, niamey$tot_cases, type="b", xlab="Week", ylab="Incidence") 
par(new=T)

plot(niamey$absweek, niamey$cum_cases, type="l", col="red", axes=FALSE, xlab=NA, ylab=NA, log="y")

axis(side = 4)
mtext(side = 4, line = 4, "Cumulative incidence") 
legend("topleft", legend=c("Cases", "Cumulative"), lty=c(1,1), pch=c(1,NA), col=c("black", "red"))
```

We calculate R0 using the apparent log-linear property of the data. Based on the initial exponential increase rate, we can then derive an expression for $R_0$ which we use in the next code block: $$R_0 = Vr + 1$$. In this case, the serial interval $V$ is between 10 and 12 days

```{r}
fit <- lm(log(cum_cases) ~ absweek, subset=absweek < 7, data = niamey)
r = fit$coefficients["absweek"]
V = c(10/7,13/7)
V*r + 1
```

## Maximum Likelihood: The Chain-Binomial Model

The Chain-binomial model was originally proposed by Bailey. It is a discrete-time stochastic alternative version to the continuous-time deterministic SIR model.

We are going to say that the probability that any given susceptible escapes contact during a period $t = 1$ is
$$1 - \exp{-\beta I /N}$$
Within the chain binomial model, the number of people who become infected during a single $t=1$ time step is governed by this probability distribution
$$I_{t+1} \sim Binomial\left(n = S_t, p = 1 - \exp{-\beta I_t/N} \right)$$
meaning that the number of susceptibles at time $t+1$ is found using the cumulative number of infecteds 
$$ S_{t + 1} = S_t - I_{t+1} = S_0 - \sum_{1}^{t+1} I_t $$

We can now think of the likelihood of getting $I_{t+1}$ people removed from the pool of susceptibles in terms of this probability distribution. The `dbinom` function takes 3 arguments: 
1. $I_{t+1}$, or the number that was removed
2. $n = S_t$, or the number of susceptibles at $t$
3. $p = 1 - \exp{-\beta I_t/N}$, or the probability of removing one susceptible
And we can calculate the sum of the log likelihoods as follows:
```{r}
llik.cb <- function(S0, beta, I){
  T = length(I) # number of time steps
  S = floor(S0 - cumsum(I[-T])) # number of susceptibles over time
  p = 1 - exp(-beta*(I[-T])/S0)
  L = - sum(dbinom(I[-1], S, p, log = TRUE))
  return(L)
}
```

Have to do some data processing; aggregate data into 2-week intervals
```{r}
 # a way of adding together weekly data in pairs
twoweek = rep(1:15, each = 2)
y = sapply(split(niamey$cases_1[1:30], twoweek), sum)
sum(y)

```

We are going to use our maximum likelihood method to estimate $S_0$ and $\beta$ together, but first let's just assume $S_0 = 6500$ and search values of $\beta$ to find a maximum likelihood:
```{r}
S0.init <- 6500
beta.cand <- seq(0,10,by = .1)
ll.vec <- rep(NA, length(beta.cand))

for (i in 1:length(beta.cand)){
  ll.vec[[i]] = llik.cb(S0 = S0.init, beta = beta.cand[[i]], I = y)
}

plot(ll.vec ~ beta.cand, ylab = "Neg Log Lik", xlab = expression(beta))
```
Find the value of beta which maximizes the log likelihood, or minimizes the negative log likelihood
```{r}
beta.cand[which.min(ll.vec)]
```

Assume that the guess of $\beta = 2.3$ is right, and search for the optimum value of $S_0$:
```{r}
beta.init <- 2.3
S0.cand <-seq(6000,9000, length=101)

ll.vec <- rep(NA, length(S0.cand))

for (i in 1:length(S0.cand)){
  ll.vec[[i]] = llik.cb(S0 = S0.cand[[i]], beta = beta.init, I = y)
}

plot(ll.vec ~ S0.cand, ylab = "Neg Log Lik", xlab = expression(S0))
```
```{r}
S0.cand[which.min(ll.vec)]
```

Now our goal is to vary both parameters simultaneously; we can use `optim` or `mle-2` from the `bbmle` package
```{r}
library(bbmle)


fit=mle2(llik.cb, # negative log likelihood function
         start=list(S0=7085, beta=2.3),  # starting conditions
         method="Nelder-Mead", # some kind of method...
         data = list(I = y))  # the data we are fitting to

summary(fit)
```

Our maximum likelihood estimates of our free parameters are $\beta = 1.89$ and $S_0 = 7815$. And we can easily obtain confidence intervals:
```{r}
confint(fit)
```

We can also get some correlations out of this fit:
```{r}
cov2cor(vcov(fit))
```

We can now use these parameters to simulate forward:
```{r}
sim.cb=function(S0, beta, I0){ 
  I=I0
  S=S0
  i=1 
  while(!any(I==0)){
    i=i+1
    I[i]=rbinom(1, size=S[i-1], prob=1- exp(-beta*I[i-1]/S0)) 
    S[i]=S[i-1]-I[i]
    }
  out=data.frame(S=S, I=I)
  return(out) 
}
```

```{r}
plot(y, type="n", xlim=c(1,18), ylab="Predicted/observed", xlab="Week")
for(i in 1:100){
  sim=sim.cb(S0=floor(coef(fit)["S0"]), beta=coef(fit)["beta"], I0=11)
  lines(sim$I, col=grey(.5))
}
points(y, type="b", col=2)
```

